---
title: "A predictive model for runs scored in a baseball game"
author: "Geoff Cooper [ geoffrey.cooper@ryerson.ca ]"
date: "July 7, 2020"
output:
  html_document:
    df_print: paged
subtitle: CKME136 - Initial Results
---
### Description
Initial results for a model containing all features proposed for the final model.
Training set is composed of game-by-game WHIP, DER and OPS, Test set is season to date average WHIP, DER and OPS.

### load packages
```{r}
# install.packages("ggplot2")
# install.packages("tidyverse")
# install.packages("dplyr")
# install.packages('Boruta')
```

### set working directory
```{r}
# setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
# getwd() 
```

### open tor_games raw data
```{r}
tor_games = read.csv("tor_games_20200707.csv")  # read csv file 
tor_games
```

## 1. Create datasets

### create Training dataset - game-by-game WHIP, DER and OPS
```{r}
tor_games_training <- data.frame(tor_games$tor_runs,tor_games$game_park_factor,tor_games$tor_home,tor_games$opp_pitcher_whip_game,tor_games$opp_team_der_game,tor_games$drurb001_ops_game,tor_games$bichb001_ops_game,tor_games$biggc002_ops_game,tor_games$davij007_ops_game,tor_games$fishd001_ops_game,tor_games$galvf001_ops_game,tor_games$gricr001_ops_game,tor_games$guerv002_ops_game,tor_games$hernt002_ops_game,tor_games$gurrl001_ops_game,tor_games$jansd001_ops_game,tor_games$maill001_ops_game,tor_games$mcgur002_ops_game,tor_games$mckib001_ops_game,tor_games$pillk001_ops_game,tor_games$smoaj001_ops_game,tor_games$sogae001_ops_game,tor_games$tellr001_ops_game)
names(tor_games_training) <- c('tor_runs','game_park_factor','tor_home','opp_pitcher_whip','opp_team_der','drurb001_ops','bichb001_ops','biggc002_ops','davij007_ops','fishd001_ops','galvf001_ops','gricr001_ops','guerv002_ops','hernt002_ops','gurrl001_ops','jansd001_ops','maill001_ops','mcgur002_ops','mckib001_ops','pillk001_ops','smoaj001_ops','sogae001_ops','tellr001_ops')
tor_games_training
```

### create Test dataset - season to date average WHIP, DER and OPS, last 150 games
```{r}
library(dplyr)
tor_games_test <- data.frame(tor_games$game_date,tor_games$tor_runs,tor_games$game_park_factor,tor_games$tor_home,tor_games$opp_pitcher_whip_todate,tor_games$opp_team_der_todate,tor_games$drurb001_ops_todate,tor_games$bichb001_ops_todate,tor_games$biggc002_ops_todate,tor_games$davij007_ops_todate,tor_games$fishd001_ops_todate,tor_games$galvf001_ops_todate,tor_games$gricr001_ops_todate,tor_games$guerv002_ops_todate,tor_games$hernt002_ops_todate,tor_games$gurrl001_ops_todate,tor_games$jansd001_ops_todate,tor_games$maill001_ops_todate,tor_games$mcgur002_ops_todate,tor_games$mckib001_ops_todate,tor_games$pillk001_ops_todate,tor_games$smoaj001_ops_todate,tor_games$sogae001_ops_todate,tor_games$tellr001_ops_todate)
names(tor_games_test) <- c('game_date','tor_runs','game_park_factor','tor_home','opp_pitcher_whip','opp_team_der','drurb001_ops','bichb001_ops','biggc002_ops','davij007_ops','fishd001_ops','galvf001_ops','gricr001_ops','guerv002_ops','hernt002_ops','gurrl001_ops','jansd001_ops','maill001_ops','mcgur002_ops','mckib001_ops','pillk001_ops','smoaj001_ops','sogae001_ops','tellr001_ops')
tor_games_test <- tor_games_test %>% filter(game_date > 20190409)
tor_games_test = subset(tor_games_test, select = -c(game_date))
tor_games_test
```

## 2. Train model - baseline

### train regression model - WHIP, DER, OPS game values
```{r}
model_fit <- lm(tor_runs~game_park_factor+tor_home+opp_pitcher_whip+opp_team_der+drurb001_ops+bichb001_ops+biggc002_ops+davij007_ops+fishd001_ops+galvf001_ops+gricr001_ops+guerv002_ops+hernt002_ops+gurrl001_ops+jansd001_ops+maill001_ops+mcgur002_ops+mckib001_ops+pillk001_ops+smoaj001_ops+sogae001_ops+tellr001_ops, tor_games_training)
summary(model_fit)
```

### evaluate model on Training data
```{r}
model_predict_training <- predict(model_fit, tor_games_training)
actuals_preds_training <- data.frame(cbind(actuals=tor_games_training$tor_runs, predicteds=model_predict_training))
correlation_accuracy_training <- cor(actuals_preds_training)
```
#### correlation of runs/game, Training data, actuals vs predicteds
```{r}
correlation_accuracy_training
actuals_preds_training
```

### histogram of runs/game, Training data, actuals vs predicteds
```{r}
library(ggplot2)
actuals <- data.frame(tor_games_training$tor_runs)
actuals$dataset <- 'Actual'
names(actuals) <- c('Runs','Dataset')
predicteds <- data.frame(model_predict_training)
predicteds$dataset <- 'Predicted'
names(predicteds) <- c('Runs','Dataset')
runs <- rbind(actuals, predicteds)
ggplot(runs, aes(x=Runs, y=..count.., color=Dataset, fill=Dataset)) + geom_histogram(binwidth=1, alpha=0.5, position="dodge") + geom_density(alpha=0) + scale_fill_manual(values=c("darkblue", "lightblue")) + 
scale_color_manual(values=c("darkblue", "lightblue")) + labs(title="Histogram of Runs/Game by Toronto Blue Jays (2019 Season)\nActuals vs Training Set results",x="Runs", y="Games") + theme_classic()
```

### plot of runs/game, Training data, actuals vs predicteds
```{r}
plot(actuals_preds_training$predicteds, actuals_preds_training$actuals, main="Plot of runs/game - Training data", xlab="Predicted", ylab="Actual")
abline(lm(actuals_preds_training$actuals~actuals_preds_training$predicteds), col="red")
```

## 3. Evaluate model - baseline

### evaluate model on Test data
```{r}
model_predict_test <- predict(model_fit, tor_games_test)
actuals_preds_test <- data.frame(cbind(actuals=tor_games_test$tor_runs, predicteds=model_predict_test))
correlation_accuracy_test <- cor(actuals_preds_test)
```
#### correlation of runs/game, Test data, actuals vs predicteds
```{r}
correlation_accuracy_test
actuals_preds_test
```
#### Means of actuals, predicteds
```{r}
mean(actuals_preds_test$actuals)
mean(actuals_preds_test$predicteds)
```
#### Normality of actuals, predicteds
```{r}
shapiro.test(actuals_preds_test$actuals)
shapiro.test(actuals_preds_test$predicteds)
```

### histogram of runs/game, Test set, actuals vs predicteds
```{r}
actuals <- data.frame(tor_games_test$tor_runs)
actuals$dataset <- 'Actual'
names(actuals) <- c('Runs','Dataset')
predicteds <- data.frame(model_predict_test)
predicteds$dataset <- 'Predicted'
names(predicteds) <- c('Runs','Dataset')
runs <- rbind(actuals, predicteds)
ggplot(runs, aes(x=Runs, y=..count.., color=Dataset, fill=Dataset)) + geom_histogram(binwidth=1, alpha=0.5, position="dodge") + geom_density(alpha=0) + scale_fill_manual(values=c("darkblue", "lightblue")) + 
scale_color_manual(values=c("darkblue", "lightblue")) + labs(title="Histogram of Runs/Game by Toronto Blue Jays (2019 Season)\nActuals vs Test Set results - baseline model",x="Runs", y="Games") + theme_classic()
```

### plot of runs/game Test data, actuals vs predicteds
```{r}
plot(actuals_preds_test$predicteds, actuals_preds_test$actuals, main="Plot of runs/game - Test data - baseline model", xlab="Predicted", ylab="Actual")
abline(lm(actuals_preds_test$actuals~actuals_preds_test$predicteds), col="red")
```

## 4. Revise model

### revise model using feature selection
```{r}
library(Boruta)
boruta_output <- Boruta(tor_runs~., data=tor_games_training, doTrace=0, maxRuns=500)
boruta_output
plot(boruta_output, cex.axis=.7, las=2, xlab="", main="Variable Importance")
boruta_table <- attStats(boruta_output)
boruta_table
```

### remove outlier, one game where runs > 13
```{r}
tor_games_training <- tor_games_training %>% filter(tor_runs < 13)
tor_games_test <- tor_games_test %>% filter(tor_runs < 13)
```

## 5. Train model - revised

### train revised regression model - use only important WHIP, DER, OPS game values
```{r}
model_fit <- lm(tor_runs~opp_pitcher_whip+opp_team_der+biggc002_ops+gricr001_ops+guerv002_ops+hernt002_ops+gurrl001_ops+jansd001_ops, tor_games_training)
summary(model_fit)
```

## 6. Evaluate model - revised

### use revised model on Test data - WHIP, DER, OPS todate average values
```{r}
model_predict_test <- predict(model_fit, tor_games_test)
actuals_preds_test <- data.frame(cbind(actuals=tor_games_test$tor_runs, predicteds=model_predict_test))
correlation_accuracy_test <- cor(actuals_preds_test)
```
#### correlation, Test data, actuals vs predicteds
```{r}
correlation_accuracy_test
actuals_preds_test
```

#### Means of actuals, predicteds sets
```{r}
mean(actuals_preds_test$actuals)
mean(actuals_preds_test$predicteds)
```
#### Normality of actuals, predicteds sets
```{r}
shapiro.test(actuals_preds_test$actuals)
shapiro.test(actuals_preds_test$predicteds)
```
### histogram of runs/game, Test set, actuals vs predicteds
```{r}
actuals <- data.frame(tor_games_test$tor_runs)
actuals$dataset <- 'Actual'
names(actuals) <- c('Runs','Dataset')
predicteds <- data.frame(model_predict_test)
predicteds$dataset <- 'Predicted'
names(predicteds) <- c('Runs','Dataset')
runs <- rbind(actuals, predicteds)
ggplot(runs, aes(x=Runs, y=..count.., color=Dataset, fill=Dataset)) + geom_histogram(binwidth=1, alpha=0.5, position="dodge") + geom_density(alpha=0) + scale_fill_manual(values=c("darkblue", "lightblue")) + 
scale_color_manual(values=c("darkblue", "lightblue")) + labs(title="Histogram of Runs/Game by Toronto Blue Jays (2019 Season)\nActuals vs Test Set results - revised model",x="Runs", y="Games") + theme_classic()
```

### plot of runs/game Test data, actuals vs predicteds
```{r}
plot(actuals_preds_test$predicteds, actuals_preds_test$actuals, main="Plot of runs/game - Test data - revised model", xlab="Predicted", ylab="Actual")
abline(lm(actuals_preds_test$actuals~actuals_preds_test$predicteds), col="red")
```